### 2022-10-08 AISTATS2023/paper.tex
updated title
Intro ready

\newcommand{\dfdx}{f^s}
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathbf{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

\aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for
  quantifying the hetegeneity of the local effects}

@@

Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations. However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE. Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty. Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are then computed from the instances that lie in each bin.  In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance. We demonstrate through synthetic and real datasets (a) the advantages of modelling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the mean effect and uncertainty.
\end{abstract}

@@

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), to understand the decision mechanism of 
%the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain the prediction of a
specific instance, whereas global methods provide an explanation that summarizes the entire model behavior. Global methods provide a universal explanation, aggregating the various local explanations into a single interpretable outcome,
usually a number or a plot. If users want to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
is vulnerable to misinterpretations. Under strong interactions and
correlations between features, the global explanation may obfuscate
heterogeneous effects that exist under the
hood~\citep{herbinger2022repid}; a phenomenon called \emph{aggregation
bias}~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category of global
explainability methods aiming at isolating the impact
of a single feature on the output\footnote{Often, FE methods also
  isolate the combined effect of a pair of features to the
  output. Combinations of more than two features are uncommon, since they are difficult to estimate and visualize.}.
FE methods suffer from aggregation bias because, often, implications of the average effect on the model behavior are unclear. For example, a feature with zero average effect may indicate no effect on the output or, 
a highly positive effect in some cases and a highly negative effect in others. 
There are
three widely-used FE methods: \emph{Partial Dependence Plots}
(PDP)\citep{friedman2001greedy}, \emph{Marignal Plots}
(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized \todo{Add a reference}for
computing erroneous effects when the input features are (highly)
correlated, which is typical in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method. However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition, ALE does not inform the user about the uncertainty, i.e.,
the level of heterogeneous effects hidden behind the average effect
due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be identified by exploring the
Individual Conditional Expectations
(ICE)~\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation using the (limited) samples of the training set,
ALE requires an additional \textit{bin-splitting} step to partition the domain of the feature of interest into consecutive non-overlapping intervals (regions) and to estimate a
constant effect in each region. The effect is estimated from the population of samples that fall
inside the region. Specifying an appropriate set of bins is of particular
importance, since ALE's interpretation is meaningful only inside each
region~\citep{molnar2022}. However, this crucial step has not raised
the appropriate attention and the approximation is vulnerable to
potential misinterpretations.\todo{I miss what exactly is the simplication of existing mehtods here. Should we add e.g., typically equisized bins?}

In this paper, we present Uncertainty-aware Accumulated Local Effects
(UALE) an extension of ALE for measuring the uncertainty of the
explanation, i.e.~how certain we are that the average explanation
would predict the feature effect to an instance drawn at random. We
show that choosing an appropriate sequence of intervals is
particularly important for a robust and unbiased estimation of the
uncertainty and we provide an algorithm that automates this step. We
also formally prove the bin-splitting conditions for an unbiased
estimation of the uncertainty. We apply our method in multiple
synthetic and real world datasets for evaluating its performance.

\paragraph{Contributions.} The contributions of this paper are:

\begin{itemize}
\item UALE, a FE method that extends ALE for quantifying the
  heterogeneous effects (uncertainty).
\item A formal proof of the bin-splitting requirements for an unbiased
  approximation of the heterogeneous effects.
\item A framework for automatically extracting regions with similar
  effects, improving the estimation and the interpretability of ALE
  plots.
\end{itemize}

We provide empirical evaluation of the method in artificial and real
datasets. The implementation of our method and the code for
reproducing all the experiments is provided in the submission and will
become publicly available upon acceptance.

>>>>

\newcommand{\dfdx}{f^s}
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathbf{X}_c}
\newcommand{\Xb}{\mathcal{X}}
\newcommand{\todo}[1]{[\textcolor{red}{#1}]}

@@

\aistatstitle{Uncertainty-aware Accumulated Local Effects (UALE) for quantifying the heterogeneity of instance-level feature effects}

@@

Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations. However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE. Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty. Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are then computed from the instances that lie in each bin.  In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance~\todo{check if this phrase explains well the concept.} We demonstrate through synthetic and real datasets (a) the advantages of modeling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the average effect and uncertainty.
\end{abstract}

@@

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), to understand the decision mechanism of complex black-box
models. XAI literature distinguishes between local and global
techniques~\citep{Molnar2020interpretable}. Local methods provide
instance-level explanations, i.e.~explain the prediction for a
specific instance, whereas global methods summarize the entire model
behavior.

Global methods create a global explanation by aggregating the
instance-level explanations into a single interpretable outcome,
usually a number or a plot. When strong interactions between features
exist, the instance-level explanations may significantly deviate from
the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
methods should quantify not only the aggregated (averaged) effect, but
also the level of heterogeneity of instance-level explanations. The
same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
of global XAI methods that isolate the impact of a single feature on
the output.\footnote{FE methods also isolate the combined effect of a
  pair of features to the output. Combinations of more than two
  features are uncommon, since they are difficult to estimate and
  visualize.} For example, a feature with zero average effect may
indicate no effect on the output or a highly positive effect in some
cases and a highly negative effect in others. The ambiguity about the
correct interpretation could be resolved, if the FE method informs
about the heterogeneity of instance-level explanations, a quantity we
refer to as the \emph{uncertainty of the effect}.

There are three widely-used FE methods: \emph{Partial Dependence
  Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
(ALE)\citep{apley2020visualizing}. ALE has been established as the
state-of-the-art for measuring the average effect, since PDP and MP
have been criticized~\citep{Gromping2020MAEP} to compute erroneous
effects when the input features are (highly) correlated. However, ALE
faces two limitations, the first concerns ALE definition and the
second ALE approximation. Regarding the definition, ALE does not
inform the user about the uncertainty. In contrast, in PDP the
heterogeneity of instance-level effects can be identified by exploring
the Individual Conditional Expectations
(ICE)~\citep{goldstein2015peeking}. Regarding the approximation, ALE
requires partitioning the feature domain into non-overlapping
intervals (bin-splitting) and estimate the effect inside each interval
(bin-effect) from the training set instances that fall inside. Finding
an appropriate sequence of bins is of particular importance, since
ALE's interpretation is meaningful only inside each
bin~\citep{molnar2022}. However, this crucial step has not raised the
appropriate attention. With the current
approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
split \emph{blindly}, i.e.~without considering the characteristics of
the instance-level effects, in \(K\) equisized bins.

In this paper, we present Uncertainty-aware Accumulated Local Effects
(UALE), a method that visualizes in a single plot both the average
feature effect and the uncertainty, i.e.~the level of heterogeneity of
instance-level effects. The average effect is defined as in ALE and
the uncertainty inherits ALE's advantages, i.e.~takes into account the
correlations between features. We also propose an accurate UALE
approximation. We formally prove that for an unbiased estimation of
the uncertainty, the partitioning of the feature domain must respect
specific conditions. Using that, we set-up an optimization problem
where we search for a partitioning, a sequence of \emph{variable-size}
intervals, (a) with enough instances (accurate estimation) and (b)
that respect the above conditions (unbiased estimation). Finally, we
propose a method for automatically finding the optimal
partitioning. We apply our method in multiple synthetic and real world
datasets for evaluating its performance. The contributions of this
paper are:

\begin{itemize}
\item UALE, a feature effect method that quantifies both the average
  effect and the uncertainty, i.e.,~the heterogeneity of instance level
  effects.
\item A formal proof of the conditions the feature domain partitioning
  must fulfill for an unbiased approximation of the uncertainty.
\item An accurate UALE approximation by automatically partitioning
  into variable-size intervals that respect the above conditions.
\end{itemize}

The implementation of our method and the code for
reproducing all the experiments is provided in the submission and will
become publicly available upon acceptance.
